### 2023-03-30 doc/_bayes_freq.qmd
save unstaged changes

Writing Bayes' theorem for one of these independent terms results in
$$
\pi(\underline{f}_t \mid \underline{r}_t) \propto
\pi(\underline{r}_t \mid \underline{f}_t)
\pi(\underline{f}_t).
$${#eq-bayes_freq}
To model the prior uncertainty in the value of the frequencies, we assume a very
simple model. Notice that the frequencies at any time point must satisfy the
normalization constraint
$$
\sum_{b = 1}^B f_t^{(b)} = 1, \;\; \forall \; f_t^{(b)} \in [0, 1].
$${#eq-freq_norm}
This means that the probability distribution of the $B$ frequency values lies on
the $B$-simplex, i.e., the $B$ dimensional space constrained such that the sum
of all the dimensions must add up to one. A natural selection for a probability
that satisfies the constraint in @eq-freq_norm is the so-called Dirichlet 
distribution. The Dirichlet distribution is parametrized by a vector
$$
\underline{\alpha}_t = (\alpha^{(1)}_t, \cdots, \alpha^{(B)}_t)^\dagger, 
\;\; \alpha_t^{(k)} > 0 \;\; \forall \; k.
$${#eq-alpha_dir}
We then choose as a prior distribution for the barcode frequency values
$\pi(\underline{f}_t)$ a Dirichlet distribution of the form
$$
\pi(\underline{f}_t) = \frac{
    \Gamma\left(\sum_{b=1}^B\alpha_t^{(b)}\right)
}{
    \prod_{b=1}^B\Gamma(\alpha_t^{(b)})
} 
\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1},
$${#eq-dirichlet}
where $\Gamma(\cdot)$ is the gamma function. @eq-dirichlet can be much more
compactly written as
$$
\underline{f}_t \sim \operatorname{Dirichlet}(\underline{\alpha}_t),
$${#eq-freq_prior}
where the symbol "$\sim$" is read "distributed as". Again, since the ultimate
calculation will be performed numerically, we can hide all of the mathematical
details in @eq-dirichlet with the simple expression in @eq-freq_prior. We are
free to choose the value of the parameters $\underline{\alpha}_t$ according to
our prior information. A reasonable choice is to assume that before observing
any data at all, we have no reason to believe any barcode is in higher abundance
compared to any other. This might not be entirely true at the beginning of the
experiment, where we expect the neutral barcodes to be in much higher frequency
compared to the mutant barcodes. Nevertheless, we have less certainty about the
following time points, making our "uniform prior over the simplex" a reasonable
choice. Therefore, we define
$$
\alpha_t^{(b)} = 1 \;\; \forall \; b.
$${#eq-alpha_choice}

For the likelihood term $\pi(\underline{r}_t \mid \underline{f}_t)$, we assume a
very simple multinomial model. The story behind this choice goes as follows:
Given a "true" underlying barcode frequency, when the sequencing machine reads a
particular barcode, this is equivalent to having thrown a $B$-sided die, where
the face of the die indicates the barcode that is read. This die is then rolled
$R_t = \sum_b r_t^{(b)}$ times, giving the values of the reads we observe at
time $t$. Notice that this model does not account for noise that might come from
sources such as genetic drift, DNA extraction, and PCR amplification. Although
these sources of noise are important, it is not clear that we have enough
statistical power to make meaningful statements about them given our data (more
on this later on). We can then compactly write the likelihood function as
$$
\underline{r}_t \mid \underline{f}_t \sim
\operatorname{Multinomial}(R_t, \underline{f}_t).
$${#eq-freq_likelihood}

This simple Multinomial-Dirichlet model for the probability of the frequency
values is convenient because it is analytically tractable. Specifically, the
Dirichlet distribution is the so-called *conjugate prior* to the multinomial
distribution. This means the posterior distribution $\pi(\underline{f}_t \mid
\underline{r}_t)$ has a closed-form solution of the same form as the prior
$\pi(\underline{f}_t)$, but with updated parameters. In other words, we can
write (without proof) the posterior distribution as
$$
\underline{f}_t \mid \underline{r}_t \sim
\operatorname{Dirichlet}(\underline{\beta}_t),
$${#eq-freq_posterior}
where the $\underline{\alpha}_t$ parameters are updated to $\underline{\beta}$
after observing the data. These updated parameters are of the form
$$
\beta_t^{(b)} = \alpha_t^{(b)} + r_t^{(b)}.
$$
Simply put, when the likelihood function is multinomial, and the prior 
distribution is Dirichlet, the posterior distribution is also Dirichlet, where
the parameters of the posterior are the parameters of the prior plus the number
of observations of the corresponding output. This fact will become handy
as we build the entire posterior distribution for our inference as defined in
@eq-split_posterior.

>>>>

Writing Bayes' theorem for one of these independent terms results in
$$
\pi(\underline{f}_t \mid \underline{r}_t) \propto
\pi(\underline{r}_t \mid \underline{f}_t)
\pi(\underline{f}_t).
$${#eq-bayes_freq}
To model the prior uncertainty in the value of the frequencies, we assume a very
simple model. Notice that the frequencies at any time point must satisfy the
normalization constraint
$$
\sum_{b = 1}^B f_t^{(b)} = 1, \;\; \forall \; f_t^{(b)} \in [0, 1].
$${#eq-freq_norm}
This means that the probability distribution of the $B$ frequency values lies on
the $B$-simplex, i.e., the $B$ dimensional space constrained such that the sum
of all the dimensions must add up to one. A natural choice for a probability
that satisfies the constraint in @eq-freq_norm is the so-called Dirichlet
distribution. The Dirichlet distribution is parametrized by a vector
$$
\underline{\alpha}_t = (\alpha^{(1)}_t, \cdots, \alpha^{(B)}_t)^\dagger, 
\;\; \alpha_t^{(k)} > 0 \;\; \forall \; k.
$${#eq-alpha_dir}
We then choose as a prior distribution for the barcode frequency values
$\pi(\underline{f}_t)$ a Dirichlet distribution of the form
$$
\pi(\underline{f}_t) = \frac{
    \Gamma\left(\sum_{b=1}^B\alpha_t^{(b)}\right)
}{
    \prod_{b=1}^B\Gamma(\alpha_t^{(b)})
} 
\prod_{b=1}^B (f_t^{(b)})^{\alpha_t^{(b)}-1},
$${#eq-dirichlet}
where $\Gamma(\cdot)$ is the gamma function. @eq-dirichlet can be much more
compactly written as
$$
\underline{f}_t \sim \operatorname{Dirichlet}(\underline{\alpha}_t),
$${#eq-freq_prior}
where the symbol "$\sim$" is read "distributed as". Again, since the ultimate
calculation will be performed numerically, we can hide all of the mathematical
details in @eq-dirichlet with the simple expression in @eq-freq_prior. We are
free to choose the value of the parameters $\underline{\alpha}_t$ according to
our prior information. A reasonable choice is to assume that before observing
any data at all, we have no reason to believe any barcode is in higher abundance
compared to any other. This might not be entirely true at the beginning of the
experiment, where we expect the neutral barcodes to be in much higher frequency
compared to the mutant barcodes. Nevertheless, we have less certainty about the
following time points, making our "uniform prior over the simplex" a reasonable
choice. Therefore, we define
$$
\alpha_t^{(b)} = 1 \;\; \forall \; b.
$${#eq-alpha_choice}

For the likelihood term $\pi(\underline{r}_t \mid \underline{f}_t)$, we assume a
very simple multinomial model. The story behind this choice goes as follows:
Given a "true" underlying barcode frequency, when the sequencing machine reads a
particular barcode, this is equivalent to having thrown a $B$-sided die, where
the face of the die indicates the barcode that is read. This die is then rolled
$R_t = \sum_b r_t^{(b)}$ times, giving the values of the reads we observe at
time $t$. Notice that this model does not account for noise that might come from
sources such as genetic drift, DNA extraction, and PCR amplification. Although
these sources of noise are important, it is not clear that we have enough
statistical power to make meaningful statements about them given our data (more
on this later on). We can then compactly write the likelihood function as
$$
\underline{r}_t \mid \underline{f}_t \sim
\operatorname{Multinomial}(R_t, \underline{f}_t),
$${#eq-freq_likelihood}
which again is simply read as the number of reads at time $t$ $\underline{r}_t$
given the frequencies at the same time $\underline{f}_t$ is distributed with
a multinomial distribution parametrized by $R_t$, the total number of reads, and
the frequency vector.

This simple Multinomial-Dirichlet model for the probability of the frequency
values is convenient because it is analytically tractable. Specifically, the
Dirichlet distribution is the so-called *conjugate prior* to the multinomial
distribution. This means the posterior distribution $\pi(\underline{f}_t \mid
\underline{r}_t)$ has a closed-form solution of the same form as the prior
$\pi(\underline{f}_t)$, but with updated parameters. In other words, we can
write (without proof) the posterior distribution as
$$
\underline{f}_t \mid \underline{r}_t \sim
\operatorname{Dirichlet}(\underline{\beta}_t),
$${#eq-freq_posterior}
where the $\underline{\alpha}_t$ parameters are updated to $\underline{\beta}_t$
after observing the data. These updated parameters are of the form
$$
\beta_t^{(b)} = \alpha_t^{(b)} + r_t^{(b)}.
$$
Simply put, when the likelihood function is multinomial, and the prior 
distribution is Dirichlet, the posterior distribution is also Dirichlet, where
the parameters of the posterior are the parameters of the prior plus the number
of observations of the corresponding output. This fact will become handy
as we build the entire posterior distribution for our inference as defined in
@eq-split_posterior.

### 2023-03-30 doc/_bayes_meanfit.qmd
save unstaged changes

Next, we turn our attention to the problem of determining the population mean
fitnesses $\underline{\bar{s}}_T$. First, we notice that our fitness model in
@eq-fitness does not include the value of the raw reads. They enter the
calculation indirectly through the inference of the frequency values we
developed in @sec-bayes_freq. This means that we can remove the conditioning of
the value of $\underline{\bar{s}}_T$ on the number of reads, obtaining a simpler
probability function
$$
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}, \underline{\underline{R}}
) = 
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}
).
$${#eq-meanfit_noreads}
Moreover, our fitness model does not directly explain how the population mean
fitness evolves over time. In other words, our model cannot explicitly compute
the population mean fitness at time $t+1$ from the information we have about
time $t$. Given this model limitation, we are led to assume that we must need
to infer each $\bar{s}_t$ independently. Expressing this for our inference
results in
$$
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}
) =
\prod_{t=1}^{T-1} \pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}),
$${#eq-meanfit_indep}
where we split our matrix $\underline{\underline{F}}$ for each time point and
only kept the conditioning on the relevant frequencies needed to compute the
mean fitness at time $t$.

@@

Recall that in @sec-fitness_model we emphasized that the frequencies $f_t^{(n)}$
do not represent the true frequency of a particular lineage in the population
but rather a "normalized number of cells." Therefore, it is safe to assume each
of the $N$ neutral lineages' frequencies is changing independently. Their
correlation of how increasing the frequency of one lineage will decrease the
frequency of others is already captured in the model presented in
@sec-bayes_freq. Thus, we write
$$
\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N) =
\prod_{n=1}^N \pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}).
$${#eq-meanfit_indep_lineages}

Now, we can focus on one of the terms on the right-hand side of
@eq-meanfit_indep_lineages. Writing Bayes theorem results in
$$
\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
$${#eq-meanfit_bayes}
Notice the likelihood defines the joint distribution of neutral barcode
frequencies conditioned on the population mean fitness. However, rewriting our
fitness model in @eq-fitness for a neutral lineage to leave frequencies on one
hand and fitness on the other results in
$$
\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t}.
$${#eq-fitness_ratio_neutral}
@eq-fitness_ratio_neutral implies that our fitness model only relates **the
ratio** of frequencies and not the individual values. To get around this 
complication, we define
$$
\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
$${#eq-gamma_def}
as the ratio of frequencies between two adjacent time points for any barcode
$b$. This allows us to rewrite the joint distribution
$\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)$ as
$$
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_freq_gamma}
Let us rephrase this subtle but necessary change of variables since it is a key
part of the inference problem: our series of independence assumptions lead us to
@eq-meanfit_bayes that relates the value of the population mean fitness
$\bar{s}_t$ to the frequency of a neutral barcode at times $t$ and $t+1$. 
However, as shown in @eq-fitness_ratio_neutral, our model functionally relates
the ratio of frequencies---that we defined as $\gamma_t^{(n)}$---and not the
independent frequencies to the mean fitness. Therefore, instead of writing the
joint distribution of the frequency values at times $t$ and $t+1$ conditioned on
the mean fitness for the likelihood, we write the joint distribution of the
barcode frequency at time $t$ and the ratio of the frequencies. These **must
be** equivalent joint distributions since there is a one-to-one mapping between
$\gamma_t^{(n)}$ and $f_{t+1}^{(n)}$ for a given value of $f_t^{(n)}$.

@@

Putting all the pieces we have developed in this section together results in an
inference for the population mean fitness values of the form
$$
\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}) \propto
\prod_{t=1}^{T-1} \left\{
    \prod_{n=1}^N \left[
        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
    \right]
    \pi(\bar{s}_t) \pi(\sigma_t)
\right\},
$${#eq-meanfit_full_inference}
where we have
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
$${#eq-meanfit_eq1}
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
\log\mathcal{N}(\bar{s}_t, \sigma_t),
$${#eq-meanfit_eq2}
$$
\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
$${#eq-meanfit_eq3}
and
$$
\sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
$${#eq-meanfit_eq4}
where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can
choose.

>>>>

Next, we turn our attention to the problem of determining the population mean
fitnesses $\underline{\bar{s}}_T$. First, we notice that our fitness model in
@eq-fitness does not include the value of the raw reads. They enter the
calculation indirectly through the inference of the frequency values we
developed in @sec-bayes_freq. This means that we can remove the conditioning of
the value of $\underline{\bar{s}}_T$ on the number of reads, obtaining a simpler
probability function
$$
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}, \underline{\underline{R}}
) = 
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}
).
$${#eq-meanfit_noreads}
Moreover, our fitness model does not directly explain how the population mean
fitness evolves over time. In other words, our model cannot explicitly compute
the population mean fitness at time $t+1$ from the information we have about
time $t$. Given this model limitation, we are led to assume that we must infer
each $\bar{s}_t$ independently. Expressing this for our inference
results in
$$
\pi(
    \underline{\bar{s}}_T \mid 
    \underline{\underline{F}}
) =
\prod_{t=1}^{T-1} \pi(\bar{s}_t \mid \underline{f}_t, \underline{f}_{t+1}),
$${#eq-meanfit_indep}
where we split our matrix $\underline{\underline{F}}$ for each time point and
only kept the conditioning on the relevant frequencies needed to compute the
mean fitness at time $t$.

@@

Recall that in @sec-fitness_model we emphasized that the frequencies $f_t^{(n)}$
do not represent the true frequency of a particular lineage in the population
but rather a "normalized number of cells." Therefore, it is safe to assume each
of the $N$ neutral lineages' frequencies is changing independently. The
correlation of how increasing the frequency of one lineage will decrease the
frequency of others is already captured in the model presented in
@sec-bayes_freq. Thus, we write
$$
\pi(\bar{s}_t \mid \underline{f}_t^N, \underline{f}_{t+1}^N) =
\prod_{n=1}^N \pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}).
$${#eq-meanfit_indep_lineages}

Now, we can focus on one of the terms on the right-hand side of
@eq-meanfit_indep_lineages. Writing Bayes theorem results in
$$
\pi(\bar{s}_t \mid f_t^{(n)}, f_{t+1}^{(n)}) \propto
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) \pi(\bar{s}_t).
$${#eq-meanfit_bayes}
Notice the likelihood defines the joint distribution of neutral barcode
frequencies conditioned on the population mean fitness. However, rewriting our
fitness model in @eq-fitness for a neutral lineage to leave frequencies on one
side and fitness on the other results in
$$
\frac{f_{t+1}^{(n)}}{f_t^{(n)}} = \mathrm{e}^{- \bar{s}_t}\tau.
$${#eq-fitness_ratio_neutral}
@eq-fitness_ratio_neutral implies that our fitness model only relates **the
ratio** of frequencies and not the individual values. To get around this 
complication, we define
$$
\gamma_t^{(b)} \equiv \frac{f_{t+1}^{(b)}}{f_t^{(b)}},
$${#eq-gamma_def}
as the ratio of frequencies between two adjacent time points for any barcode
$b$. This allows us to rewrite the joint distribution
$\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t)$ as
$$
\pi(f_t^{(n)}, f_{t+1}^{(n)} \mid \bar{s}_t) =
\pi(f_t^{(n)}, \gamma_{t}^{(n)} \mid \bar{s}_t).
$${#eq-joint_freq_gamma}
Let us rephrase this subtle but necessary change of variables since it is a key
part of the inference problem: our series of independence assumptions lead us to
@eq-meanfit_bayes that relates the value of the population mean fitness
$\bar{s}_t$ to the frequency of a neutral barcode at times $t$ and $t+1$. 
However, as shown in @eq-fitness_ratio_neutral, our model functionally relates
the ratio of frequencies---that we defined as $\gamma_t^{(n)}$---and not the
independent frequencies to the mean fitness. Therefore, instead of writing for
the likelihood the joint distribution of the frequency values at times $t$ and
$t+1$ conditioned on the mean fitness, we write the joint distribution of the
barcode frequency at time $t$ and the ratio of the frequencies. These **must
be** equivalent joint distributions since there is a one-to-one mapping between
$\gamma_t^{(n)}$ and $f_{t+1}^{(n)}$ for a given value of $f_t^{(n)}$. Another
way to phrase this is to say that knowing the frequency at time $t$ and at time
$t+1$ provides the same amount of information as knowing the frequency at time
$t$ and the ratio of the frequencies. This is because if we want to obtain
$f_{t+1}^{(n)}$ given this information, we simply compute
$$
f_{t+1}^{(n)} = \gamma_t^{(n)} f_t^{(n)}.
$${#eq-meanfit_ft+1_gamma}

@@

Putting all the pieces we have developed in this section together results in an
inference for the population mean fitness values of the form
$$
\pi(
    \underline{\bar{s}}_T, \underline{\sigma}_T \mid \underline{\underline{F}}
) \propto
\prod_{t=1}^{T-1} \left\{
    \prod_{n=1}^N \left[
        \pi(f_t^{(n)} \mid \gamma_t^{(n)}) 
        \pi(\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t)
    \right]
    \pi(\bar{s}_t) \pi(\sigma_t)
\right\},
$${#eq-meanfit_full_inference}
where we have
$$
f_t^{(n)} \mid \gamma_t^{(n)} \sim 
\operatorname{Uniform} \left(0, \frac{1}{\gamma_t^{(n)}} \right),
$${#eq-meanfit_eq1}
$$
\gamma_t^{(n)} \mid \bar{s}_t, \sigma_t \sim 
\log\mathcal{N}(\bar{s}_t, \sigma_t),
$${#eq-meanfit_eq2}
$$
\bar{s}_t \sim \mathcal{N}(0, \sigma_{\bar{s}_t}),
$${#eq-meanfit_eq3}
and
$$
\sigma_t \sim \operatorname{Half}-\mathcal{N}(0, \sigma_{\sigma_t}),
$${#eq-meanfit_eq4}
where $\sigma_{\bar{s}_t}$ and $\sigma_{\sigma_t}$ are free parameters we can
choose.

### 2023-03-30 doc/_fitness_model.qmd
save unstaged changes

1. Uncertainty in the determination of frequencies. Our model relates
frequencies between adjacent time points to the fitness of the corresponding
strain. However, we do not measure frequencies directly. Instead, our data
consists of a vector of barcode counts $\underline{r}_t$ for each of the $T$
time points in which the measurements were taken.
2. Uncertainty in the value of the population mean fitness. As we will see later
on, we use neutral lineages barcodes, which, by definition have mean fitness
$s^{(n)}$ to determine the value of the population mean fitness $\bar{s}_t$.
This inference will come with its own uncertainty that must be propagated to the
value of the relative fitness for the mutants.
3. Uncertainty in each of the mutant's fitness values.

>>>>

1. Uncertainty in the determination of frequencies. Our model relates
frequencies between adjacent time points to the fitness of the corresponding
strain. However, we do not measure frequencies directly. Instead, our data
consists of a vector of barcode counts $\underline{r}_t$ for each of the $T$
time points in which the measurements were taken.
2. Uncertainty in the value of the population mean fitness. As we will see later
on, we use neutral lineages barcodes, which, by definition have mean fitness
$s^{(n)} = 0$ to determine the value of the population mean fitness $\bar{s}_t$.
This inference will come with its own uncertainty that must be propagated to the
value of the relative fitness for the mutants.
3. Uncertainty in each of the mutant's fitness values.
